Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-14 Paper/app.tex
..

For the high-dimensional case, we aim to investigate the impact of the uncertain
pre-exponents ($A_i$'s) as well as the activation energies ($E_{a,i}$'s) on ignition 
delay during the H$_2$/O$_2$ reaction. The $A_i$'s are considered to be uniformly
distributed in the interval, $[0.9A_i^\ast, 1.1A_i^\ast]$. The $E_{a,i}$'s for all
reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$
are considered to be uncertain and uniformly distributed in the interval: 
$[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$. The nominal values, $A_i^\ast$ and $E_{a,i}^\ast$
corresponding to the different reaction rates are provided in~\cite{Yetter:1991}.

The gradient-free approach discussed earlier in~\ref{sub:gradfree} was used to compute the
active subspace. Using the iterative procedure, the convergence of the eigenvectors
was examined by tracking $\delta \hat{\mat{W}}_{1,j}^{(i)}$. In Figure~\ref{fig:conv_app} (left),
components of the converged eigenvector based on 50 samples using $\tau$ = 0.2 are plotted.
To illustrate convergence, we compare them with components of the eigenvector obtained
using 100 samples (twice as many). The variation of maximum $\delta w_{1,j}^{(i)}$ between
successive iterations during the course of the iterative procedure is illustrated in 
Figure~\ref{fig:conv_app} (right). 
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv10}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  
$\delta \hat{\mat{W}}_{1,j}^{(i)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:conv_app}
\end{center}
\end{figure}
%
The dominant eigenvector is observed to converge with only 50 samples in the 33-dimensional space which
is remarkably efficient.

In Figure~\ref{fig:hd}, we illustrate the resulting eigenvalue spectrum and the SSP.
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: SSP of the computed active subspace, and the
corresponding 1-D linear fit.} 
\label{fig:hd}
\end{center}
\end{figure}
%
The second eigenvalue is observed to be roughly four orders of magnitude smaller than the first 
indicating the existence of a 1-dimensional active subspace. This is further confirmed by the SSP
 which shows that the variability in $G$ can be approximated using a 1-dimensional regression
fit ($\tilde{G}$) which could be used as a surrogate. A linear, quadratic, and a cubic fit to the SSP was performed.
The linear fit illustrated in Figure~\ref{fig:hd} (right) was found to be the most accurate. In~\ref{sub:verify},
we investigate the accuracy of the 1D surrogate based on the linear fit by estimating the relative L-2
norm of its discrepancy with model evaluations at an independent set of samples, and by comparing the
respective probability distributions.

Components of the dominant eigenvector in Figure~\ref{fig:conv_app} (left) are used for estimating the
activity scores, plotted below in Figure~\ref{fig:as_33D}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/as_A_33D}
    \includegraphics[width=0.45\textwidth]{./Figures/as_E_33D}
\caption{A bar-graph illustrating individual activity scores for the uncertain $A_i$' (left) and $E_{a,i}$'s (right).}
\label{fig:as_33D}
\end{center}
\end{figure}
%
From the above plot, it can be inferred that the ignition delay is predominantly sensitivity towards $A_1$, $A_9$, and
$E_{a,1}$. Furthermore, it is moderately sensitivity towards $A_{15}$ and $A_{17}$ while sensitivity towards remaining
inputs is found to be low or negligible. Note that the sensitivity results for the pre-exponents is found to be consistent
with the 19-dimensional case presented in~\ref{sec:method}. This observation can also be exploited for reducing the 
dimensionality of the problem from 33 to 3-5 depending upon the required level of accuracy in terms of characterizing 
the uncertainty in the model output. The sensitivity-driven dimension reduction for uncertainty quantification has been
discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. We verify
the accuracy of the 1-d regression fit to $G(\hat{\mat{W}}_1^\top\bm{\xi})$ in the following section.

>>>>

For the high-dimensional case, we aim to investigate the impact of the uncertain
pre-exponents ($A_i$'s) as well as the activation energies ($E_{a,i}$'s) on ignition 
delay during the H$_2$/O$_2$ reaction. The $A_i$'s were considered to be uniformly
distributed in the interval, $[0.9A_i^\ast, 1.1A_i^\ast]$. The $E_{a,i}$'s for all
reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$ (due to 
zero nominal values for $E_a$)
were considered to be uncertain and uniformly distributed in the interval: 
$[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$. The nominal values, $A_i^\ast$ and $E_{a,i}^\ast$
corresponding to the different reaction rates are provided in~\cite{Yetter:1991}.

The gradient-based approach discussed earlier in~\ref{sub:grad} was used to compute the
active subspace. Using the iterative procedure, the convergence of the eigenvectors
was examined by tracking $\max(\delta \hat{\mat{W}}_{1,j}^{(i)})$, plotted in 
Figure~\ref{fig:conv_app} (right). In Figure~\ref{fig:conv_app} (left), individual
components of the converged eigenvector are illustrated. The convergence was established using
a $\tau$ value of 0.02. Finite difference was used to compute model gradients at 40 samples in
the input domain. Hence, a total of 1360 model evaluations were required for this purpose.  
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv10}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  
$\max(\delta \hat{\mat{W}}_{1,j}^{(i)})$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:conv_app}
\end{center}
\end{figure}
%

In Figure~\ref{fig:hd} (left), we illustrate the resulting eigenvalue spectrum. The second eigenvalue is
found to be roughly two orders of magnitude smaller than the first, indicating that the active subspace
is 1-dimensional. This is further confirmed by the SSP plots in Figure~\ref{fig:hd} (right). Specifically
in the case of gradient-based approach, the SSP exhibits a linear variation that is reasonably
captured by its linear fit, $\tilde{G}$. Whereas, the SSP based on the gradient-free approach
is relatively more scattered. Consequently, the linear fit exhibits a discrepancy with the corresponding
fit for the gradient-based SSP. 
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: SSP of the computed active subspace, and the
corresponding 1-D linear fit.} 
\label{fig:hd}
\end{center}
\end{figure}
%
We plot the normalized activity scores as well as the total Sobol' indices for the 33 uncertain 
rate-controlling parameters in Figure~\ref{fig:as_33D}. Note that the normalized activity scores
as well as the Sobol' indices
were evaluated using respective 1-dimensional surrogates for the two approaches.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/as_33D_new}
\caption{A bar-graph illustrating individual activity scores for the uncertain $A_i$' (left) and $E_{a,i}$'s (right).}
\label{fig:as_33D}
\end{center}
\end{figure}
%

Several useful inferences can be drawn from the above plot. Firstly, the total Sobol indices are found to be
consistent
with the normalized activity scores evaluated uing the gradient-based approach. Whereas, the gradient-free
approach yields consistent results for the $A_i$'s, it does not capture the sensitivity associated with
$E_{a,15}$. While this observation bolsters our confidence in the gradient-based approach, it clearly
underscores the shortcoming of the gradient-free approach. Numerical errors are incurred when approximating
the model gradients using the local linear approximation in this case. In~\ref{sub:verify}, we assess the
suitability of the gradient-free approach in quantifying the uncertainty in the ignition delay. 
Secondly, our findings indicate that the  
the ignition delay is predominantly sensitivity towards $A_1$, $A_9$, $E_{a,1}$, and $E_{a,15}$ and 
moderately sensitivity towards $A_{15}$ and $A_{17}$. Sensitivity towards the remaining uncertain inputs is
found to be low or negligible. Note that his observation can also be exploited for reducing the 
dimensionality of the problem from 33 inputs to 4-6 inputs depending upon the required level of accuracy.
The sensitivity-driven dimension reduction for uncertainty quantification has been
discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. Furthermore, we have
shown that the dominant eigenspace can be used to approximate global sensitivity measures with no additional effort.
The accuracy of the 1-dimensional surrogate, $\tilde{G}$, obtained using the two approaches is assessed in the following section.
